Hydras \& Co.

### 2024-01-29 doc/epsilon0-chapter.tex

In this chapter, we adapt to \coq{} the well-known~\cite{KP82}  proof that Hercules eventually wins every battle, whichever the strategy  of each player.
In other words, we present  a formal and self contained proof of termination  of all [free] hydra battles.
First, we take from Manolios and Vroon~\cite{Manolios2005} a representation of the ordinal $\epsilon_0$ as terms in Cantor normal form. Then, we define a variant for hydra battles as a measure that maps any hydra to some ordinal strictly less than $\epsilon_0$.

@@

The ordinal \(\epsilon_0\) is the least ordinal number that satisfies 
the equation \(\alpha = \omega^\alpha\), where \(\omega\) is 
the least infinite ordinal.
Thus, we can intuitively consider \(\epsilon_0\) as an
\emph{infinite} \(\omega\)-tower.

@@

Any ordinal strictly less that \(\epsilon_0\) 
can be finitely represented by a unique  \emph{Cantor normal form}, 
that is, an expression  which is either  the ordinal \(0\) or 
a sum  \(\omega^{\alpha_1} \times n_1 + \omega^{\alpha_2} \times n_2 + 
  \dots + \omega^{\alpha_p} \times n_p\) where all the \(\alpha_i\) 
are ordinals in Cantor  normal form, \(\alpha_1 > \alpha_2 > \alpha_p\), 
and all the \(n_i\) are positive integers.

@@

In the rest of this section, we define an inductive type for representing in \texttt{Coq}
all the ordinals strictly  less than  \(\epsilon_0\), then extend some arithmetic operations
to this type, and finally prove that our representation fits well with 
the expected mathematical properties: the order we define is a well order, 
and the decomposition into Cantor normal form  is consistent 
with the implementation of the arithmetic operations of exponentiation of base \(\omega\) 
and addition.

\paragraph*{Remark}
\label{sec:orgheadline65}
Unless explicitly mentioned, the term ``ordinal" will be used instead of
``ordinal strictly less than \(\epsilon_0\)" (except in Chapter~\ref{chap:schutte} where it stands for ``countable ordinal'').

@@

\paragraph{Remark}
For simplicity's sake, we chose to forbid  expressions of the form $\omega^\alpha\times 0 + \beta$. Thus, the construction (\texttt{cons $\alpha$ $n$ $\beta$}) is intended to represent the
ordinal $\omega^\alpha\times(n+1)+\beta$ and not $\omega^\alpha\times n+\beta$.
In a future version, we should replace  the type \texttt{nat} with \texttt{positive} in \texttt{T1}'s 
definition. But this replacement would take a lot of time \dots{}

@@

Note that \texttt{omega} is not an identifier in \HydrasLib, thus any tactic like \texttt{unfold omega} would fail.

\index{gaiabridge}{The ordinal $\omega$}
\paragraph*{\gaiasign}
In \texttt{gaia.ssete9}, the ordinal $\omega$ is bound to the \emph{constant}  \texttt{T1omega}.

@@

\index{hydras}{Projects}
\begin{project}
Design  (in \ocaml?) a set of tools for systematically pretty printing ordinal terms in Cantor normal form.
\end{project}

@@

Links between the function \texttt{compare} and the relations
\texttt{lt} and \texttt{eq} are established through the following lemmas (~\vref{sect:comparable-def}).
\vspace{4pt}

>>>>

In this chapter, we adapt to \coq{} the well-known proof~\cite{KP82} that Hercules eventually wins every battle, whichever the strategy  of each player.
In other words, we present  a formal and self-contained proof of termination  of all [free] hydra battles.
First, we take from Manolios and Vroon~\cite{Manolios2005} a representation of the ordinal $\epsilon_0$ as terms in Cantor normal form. Then, we define a variant for hydra battles as a measure that maps any hydra to some ordinal strictly less than $\epsilon_0$.

@@

The ordinal \(\epsilon_0\) is the least ordinal number that satisfies 
the equation \(\alpha = \omega^\alpha\), where \(\omega\) is 
the least infinite ordinal\footnote{For a precise ---\,\emph{i.e.} mathematical\,--- definition of $\omega^\alpha$, please see Sect.~\vref{sect:AP-and-phi0}.} .
Thus, we can intuitively consider \(\epsilon_0\) as an
\emph{infinite} \(\omega\)-tower.

@@

Any ordinal strictly less that \(\epsilon_0\) 
can be finitely represented by a unique  \emph{Cantor normal form}, 
that is, an expression  which is 
a sum  \(\omega^{\alpha_1} \times n_1 + \omega^{\alpha_2} \times n_2 + 
  \dots + \omega^{\alpha_p} \times n_p\) where $p\in\mathbb{N}$, all the \(\alpha_i\) 
are ordinals in Cantor  normal form, \(\alpha_1 > \alpha_2 > \alpha_p\)
and all the \(n_i\) are positive integers.

@@

In the rest of this section, we define an inductive type for representing in \texttt{Coq}
all the ordinals strictly  less than  \(\epsilon_0\), then extend some arithmetic operations
to this type, and finally prove that our representation fits well with 
the expected mathematical properties: the order we define is a well order, 
and the decomposition into Cantor normal form  is consistent 
with other usual definitions of ordinals, for instance in \gaia~\cite{Gaia}, Schütte's book~\cite{schutte}, or larger ordinal notations cf. Chapter~\vref{chap:gamma0}.

\paragraph*{Remark}
\label{sec:orgheadline65}
Unless explicitly mentioned, the term ``ordinal" will be used instead of
``ordinal strictly less than \(\epsilon_0\)" (except in Chapter~\ref{chap:schutte} where it stands for ``countable ordinal'', and Chapter~\ref{chap:gamma0}.).

@@

\paragraph{Remark}
For simplicity's sake, we chose to forbid  expressions of the form $\omega^\alpha\times 0 + \beta$. Thus, the construction (\texttt{cons $\alpha$ $n$ $\beta$}) is intended to represent the
ordinal $\omega^\alpha\times(n+1)+\beta$ and not $\omega^\alpha\times n+\beta$.
In a future version, we would like to replace  the type \texttt{nat} with standard library's type \texttt{positive} in \texttt{T1}'s 
definition. But this replacement would take a lot of time \dots{}

@@

Note that \texttt{T1omega} is not an identifier in \HydrasLib, thus any tactic like \texttt{unfold T1omega} would fail.

\index{gaiabridge}{The ordinal $\omega$}
\paragraph*{\gaiasign}
In \texttt{gaia.ssete9}, the ordinal $\omega$ is bound to the \emph{constant}  \texttt{T1omega} (not a notation).

@@

\index{hydras}{Projects}
\begin{project}
Design tools for systematically pretty printing ordinal terms in Cantor normal form.
\end{project}

@@

% Links between the function \texttt{compare} and the relations
% \texttt{lt} and \texttt{eq} are established through the following lemmas (~\vref{sect:comparable-def}).
% \vspace{4pt}

### 2024-01-29 doc/gaia-chapter.tex
small improvements
minor corrections

Functions \texttt{h2g} and \texttt{g2h} allow us to define
a notion of ``data-refinement''  for constants, functions, predicates and relations. The following definitions express that some
constant, function, relation defined in \HydrasLib ``implements'' the same concept of \gaia.

@@

\inputsnippets{T1Bridge/refines1R, T1Bridge/refines2R}

@@

The following command lists us `gaia's lemmas (thanks to
\gaia's  \texttt{cantor\_scope}.
\inputsnippets{T1Bridge/SearchDemoa}

With \texttt{t1\_scope}:
\inputsnippets{T1Bridge/SearchDemob}

@@

\section{Rapidly growing arithmetic functions}
\index{gaiabridge}{Rapidly growing functions}

\subsection{A type for well-formed ordinal terms}
\label{sect:E0-gaia-hydras}
For compatibility's sake, we add a clone of \texttt{hydra-battles}' type \texttt{E0} defined in \href{../theories/html/hydras.Epsilon0.E0.html}{Epsilon0.E0}.

@@

@@

@@

>>>>

Functions \texttt{h2g} and \texttt{g2h} allow us to define
a notion of ``data-refinement''  for constants, functions, predicates and relations. The following definitions express that a given
constant, function, relation defined in \HydrasLib ``implements'' the same concept of \gaia.

From~\href{../theories/html/gaia_hydras/T1Bridge.html}%
{\texttt{gaia\_hydras.T1Bridge}}.

@@

\inputsnippets{T1Bridge/refines1R}

\inputsnippets{T1Bridge/refines2R}

@@

The following command lists us `gaia's lemmas (thanks to
\gaia's  \texttt{cantor\_scope}).
\inputsnippets{T1Bridge/SearchDemoa}

Within \texttt{t1\_scope}:
\inputsnippets{T1Bridge/SearchDemob}

@@

\subsection{A type for well-formed ordinal terms}
\label{sect:E0-gaia-hydras}
For compatibility's sake, we add a clone of \texttt{hydra-battles}' type \texttt{E0} defined in \href{../theories/html/hydras.Epsilon0.E0.html}{Epsilon0.E0}. This type is mainly used in
the definition with \texttt{Equations} of rapidly growing functions indexed by ordinals (see Sect.~\vref{sect:gaia-rapidly-growing}).

@@

\subsubsection{Definitions}

@@

\subsubsection{Main lemmas about \texttt{E0}}

@@

\section{Rapidly growing arithmetic functions}
\index{gaiabridge}{Rapidly growing functions}
\label{sect:gaia-rapidly-growing}

### 2024-01-29 doc/hydras.tex

\subsubsection{Verbatim quotations}

Quotations of \coq{} source  from other libraries (\coq's standard library, borrowed plug-ins)  are displayed as follows.

\begin{Coqsrc}
  Inductive CompareSpec (Peq Plt Pgt : Prop) :
  comparison -> Prop :=
    CompEq : Peq -> CompareSpec Peq Plt Pgt Eq
  | CompLt : Plt -> CompareSpec Peq Plt Pgt Lt
  | CompGt : Pgt -> CompareSpec Peq Plt Pgt Gt.
\end{Coqsrc}

We use also verbatim code inclusions when the examples lead to too long computations.

@@

\begin{Coqbad}
Qed.
\end{Coqbad}

 \subsection{Remark}

@@

\part{Ackermann, G\"{o}del, Peano\,\dots}

\include{chap-intro-goedel}
\include{chapter-primrec} 
\include{chapter-fol}
\include{chapter-natural-deduction}
\include{chapter-lnn-lnt}
\include{chapter-encoding}
\include{chapter-expressible}

>>>>

\subsubsection{Verbatim quotations}
In some situations, we replace \alectr snippets with
verbatim blocks.

\begin{itemize}
\item When the quoted source belongs to some library on which we do not have the write permission, we cannot include
  directives for generating snippets.
 For instance, the following code belongs to \coq's standard library.
  \begin{Coqsrc}
  Inductive CompareSpec (Peq Plt Pgt : Prop) :
  comparison -> Prop :=
    CompEq : Peq -> CompareSpec Peq Plt Pgt Eq
  | CompLt : Plt -> CompareSpec Peq Plt Pgt Lt
  | CompGt : Pgt -> CompareSpec Peq Plt Pgt Gt.
\end{Coqsrc}

\item 
We use also verbatim code inclusions when the examples would lead to too long computations during the compilation and the
documentation generation.

@@

\begin{Coqbad}
Qed.
\end{Coqbad}
\end{itemize}
 \subsection{Remark}

@@

%\part{Ackermann, G\"{o}del, Peano\,\dots}

%\include{chap-intro-goedel}
\include{chapter-primrec} 
% \include{chapter-fol}
% \include{chapter-natural-deduction}
% \include{chapter-lnn-lnt}
% \include{chapter-encoding}
% \include{chapter-expressible}
